---
title: "Applied Logistic Regression - Exercise Week 4"
author: "Yago Durán Cid"
date: "06/06/2015"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: lualatex
---

**WEEK  4**  
  
  
*Exercise 1:*  
  
  
The data in hyponatremia.dta derive from an epidemiological study of hyponatremia (a life-threatening condition) among runners of the 2002 Boston Marathon. Hyponatremia is defined as an electrolyte disturbance in which the serum sodium concentration is lower than normal (<135 mmol/l). The aim of the study was to determine whether a runner experienced hyponatremia and to identify the principal risk factors. Participants in the 2002 Boston Marathon completed a survey including demographic and anthropometric characteristics (Body Mass Index) one or two days before the race. After the race, runners provided a blood sample in order to measure their serum sodium concentration and completed a questionnaire detailing their urine output during the race. Prerace and postrace weights were also recorded.  
a. Perform a logistic regression analysis with R using nas135 as dependent
variable and female as the only independent variable. Interpret the coefficients of
the model.  
  
```{r,echo=FALSE}
#Uploads hyponatremia in R
#We name the dataset as data
data <- read.csv("~/Coursera/Applied Logistic Regression/Homework/hyponatremia.csv")

#We use glm function in R to get estimated parameters
lrmodel<-glm(data=data,nas135~female,family="binomial")
summary(lrmodel)
```  
  
The p-value of female variable being significant is `r pchisq(lrmodel$null.deviance-lrmodel$deviance,df=1,lower.tail=FALSE)`which well below 0.05.  
Being female a binary variable the parameter $\hat{\beta_1}$ associated with the individual being female is the log of odd ratio as follows:  
$\hat{\beta_1}=log(odd\,ratio_{female=1})-log(odd\,ratio_{female=0})$ and exponentiating $e^{\hat{\beta_1}}=\frac{odd\,ratio_{female=1}}{odd\,ratio_{female=0}}$  
  
Given that in our case $\hat{\beta_1}=$ `r lrmodel$coeff[2]` the odd ratio of hyponatremia  among	females	is $e^{1.2259618}=$ `r exp(lrmodel$coeff[2])` times that of males  
 
b. Fit a model with runtime as the only independent variable. Interpret the
coefficient for runtime. c. Calculate the Odds Ratio for the variable runtime and interpret it.  
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel2<-glm(data=data,nas135~runtime,family="binomial")
summary(lrmodel2)
```  

The log odd ratio of increases by `r lrmodel2$coeff[2]` for every additional minute in run time.  
  
Given that $odd\,ratio=e^{\hat{\beta_1}}$ the increase of the odd ratio by every minute is $e^{runtime_0\hat{\beta_1}}-e^{(runtime_0+1)\hat{\beta_1}}=e^{(runtime_0-runtime_0+1)\hat{\beta_1}}=e^{\hat{\beta_1}}$  Substituting $\hat{\beta_1}$ by `r lrmodel2$coeff[2]` we get the estimate for the increase of the odds ratio of hyponatremia by every additional minute of run time is $e^{0.0155019}$ = `r exp(lrmodel2$coeff[2])`   
Moreover, $Growth\,odds=\frac{odds_1}{odds_0}-1=odds\,ratio-1=$ `r (exp(lrmodel2$coeff[2])-1)*100`% This is, by every additional munite of run time, the odds of hyponatremia increases by `r (exp(lrmodel2$coeff[2])-1)*100`%  

d. Interpret the coefficient for the constant in the model with runtime as the only
independent variable. Does it make sense? If not, what can you do to obtain a
coefficient for the constant which is easily interpreted?  
  
The constant parameter value would describe the log odds of a person who would have run the marathon in zero minutes which makes no sense at all.  
Alternatively, we centre the runtime variable in our dataset (subtracting the mean to every runtime value in the dataset) the estimated parameter for the constant would describe the log odds of a runner that finished the marathon in the average time.  

```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel3<-glm(data=data,nas135~scale(runtime,center=TRUE,scale=FALSE),family="binomial")
summary(lrmodel3)
``` 
  
This is, the odds of hyponatremia for a runner who finished the marathon in the average run time is $e^{-2.096210}=$ `r exp(lrmodel3$coeff[1])`  
  
e. Calculate the Odds Ratio of hyponatremia of a runner who takes 2 hours more
than another runner, and the corresponding 95% Confidence Interval  
  
Using the growth factor estimated above: Odds of hyponatremia for a runner that takes two hours more than references runner to finish the marathon is $e^{\hat{120\beta_1}}$ = `r exp(lrmodel2$coeff[2]*120)` times larger.  
The lower bound of 95% confidence interval for the odds is `r exp(confint.default(lrmodel2,"runtime")[1]*120)`  
The upper bound of 95% confidence interval for the odds is `r exp(confint.default(lrmodel2,"runtime")[2]*120)`  
  
f. Fit a model with female and runtime as independent variables. Interpret both
coefficients.  
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel4<-glm(data=data,nas135~female+runtime,family="binomial")
summary(lrmodel4)
```  
  
The log odd ratio of hyponatremia is `r lrmodel4$coeff[2]` for a female than for a male given that both finsih the race in the same time. This is, the odds of hyponatremia is `r (exp(lrmodel4$coeff[2])-1)*100`% higher for a female than for a male given that both finish the race in the same time.

The log odd ratio of hyponatremia increases by `r lrmodel4$coeff[3]` for every additional minute a runner takes to finish the marathon regardless of sex. This is, the odds of hyponatremia increases by `r (exp(lrmodel4$coeff[3])-1)*100`% for every additional minute a runner takes to finish the race regardless of the sex. 

g. Compare the coefficients for female in the model with female as the only
independent variable with that in the model that contains female and runtime.
What is the percentage change in the coefficient of female?  
  
The estimated parameter for female in the model excluding runtime is `r (lrmodel$coeff[2]/lrmodel4$coeff[2]-1)*100`% higher than the estimated parameter including runtime. Having such a material difference suggests that possible confounding by runtime, provided there is no interaction.  

h. Calculate the Odds Ratio of hyponatremia for a female compared to a male who
completes the marathon in the same time.  
  
Odds ratio is $e^{\hat{\beta_{female}}}$ = `r exp(lrmodel4$coeff[2])`  
  
i. What type of association do you expect between the variables female and
runtime? Answer this question before looking at the data, only on the basis of 
the observed change in the coefficient for female when runtime is entered into
the model. Then make a box-plot of runtime by female.  
  
Given that the estimated parameter for female increases when runtime is excluded and that runtime parameter is positive, we can conclude that there is a positive correlation between the runner being female and the runtime (i.e.: female runners are, on average, slower than male ones)  

```{r, echo=FALSE,fig.align='center'}

boxplot(runtime~female,data=data,pch="*",col="orange",ylab="runtime",xlab="female")
```  
  
  
j. Assess whether there is an interaction between female and runtime  
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel5<-glm(data=data,nas135~female*runtime,family="binomial")
summary(lrmodel5)
```  

The p-value for the interaction between female and runtime is 0.669123 and, therefore, interaction between sex and runtime is not significant at 95% confidence.

k. Add to the model that contains female and runtime a dichotomous variable
wgain which takes the value of 0 if wtidff ≤ 0, and the value of 1 if wtidff > 0.
Test for interaction between female and wgain.  
    
The model including the dichotomus variable wgain described above:  
  

```{r,echo=FALSE}
#We subset from the dataset the variables to build the model adding wgain
data2<-data.frame("nas135"=data$nas135,"female"=data$female
                  ,"runtime"=data$runtime,"wgain"=as.integer(data$wtdiff>0))

#We use glm function in R to get estimated parameters
lrmodel6<-glm(data=data2,nas135~female+wgain+runtime,family="binomial")
summary(lrmodel6)
```    
  
Including interaction between female and wgain:  
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel7<-glm(data=data2,nas135~female*wgain+runtime,family="binomial")
summary(lrmodel7)
```    
  
The interaction is not significant at 95% but it is at 90% confidence.  

l. On the basis of the model with the interaction term, calculate the Odds Ratios of hyponatremia for males who gain weight as compared to those who don’t.
Repeat this exercise for a female. Interpret your findings.  
  
For males, the odd ratio of those who gain weigth vs. those who doesn't is $e^{2.40100}$ = `r exp(lrmodel7$coeff[3])`  
For females, the odd ratio of those who gain weigth vs. those who doesn't is $e^{2.40100-1.20186}$ = `r exp(lrmodel7$coeff[3]+lrmodel7$coeff[5])`  
  
A male who experiences weight gain during a marathon has an odds of hyponatremia  `r exp(lrmodel7$coeff[3])` times higher than that of a male who does not gain weight. On the other hand, a female who experiences weight gain during a marathon has an odds of hyponatremia `r exp(lrmodel7$coeff[3]+lrmodel7$coeff[5])` times higher than that of a female who does not gain weight.  
  

m. Compare using the Likelihood Ratio test the model with female and runtime with
a model with female, runtime, wgain, urinat3p and bmi. (Hint: the 2 models
must be fitted on the same set of observations. Be aware of missing values in
some of these variables). How many degrees of freedom does the test statistic
have?  
  
We exclude all missing values and run the model with all variables  
  
```{r,echo=FALSE}
#We exclude from the data missing values
data3<-data.frame(data2,"urinat3p"=data$urinat3p,"bmi"=data$bmi)

#We build a dataframe showing NA values
#Since true values are 1 and false are zero, the observation with no missing
#values will be those with a sum of the row = zero
missing<-is.na(data3)
missing<-apply(missing,1,sum)

data3<-data3[missing==0,]

lrmodel8<-glm(data=data3,nas135~female+wgain+runtime+urinat3p+bmi
              ,family="binomial")
summary(lrmodel8)
```    
  
On the dataset with no missing values we run the model with female and runtime variables only:  
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel9<-glm(data=data3,nas135~female+runtime,family="binomial")
summary(lrmodel9)
```    
  
The p-value of the likelihood test (based on the deviance of both models provided by R) is `r pchisq(lrmodel9$deviance-lrmodel8$deviance,df=3,lower.tail=FALSE)` so we can confirm that the model with 5 variables is better than the one with only 2.  
  
A model including only female, runtime and wgain: 
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel10<-glm(data=data3,nas135~female+runtime+wgain,family="binomial")
summary(lrmodel10)  
```   
  
Now, the p-value of the likelihood test comparing the model with all variables and the one with three is `r pchisq(lrmodel10$deviance-lrmodel8$deviance,df=2,lower.tail=FALSE)` so we can confirm that the model without bmi and urine variables is better.  