---
title: "Applied Logistic Regression - Exercise Week 5"
author: "Yago Durán Cid"
date: "13/06/2015"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: lualatex
---

**WEEK 5**  
    
*Exercise 1:*  
  

Use the hyponatremia.dta dataset to complete the following  
a. Assess the association between hyponatremia (dichotomous variable nas135) and
sex (variable female) by making a 2 by 2 table. Calculate the odds ratio of
hyponatremia of a female compared to a male. Compute the 95% confidence interval
for this odds ratio. Interpret the findings.  
  
```{r,echo=FALSE}
#Uploads hyponatremia in R
#We name the dataset as data
data <- read.csv("~/Coursera/Applied Logistic Regression/Homework/hyponatremia.csv")

#We create the contingency table with nas135 in the columns and female in the rows
NaSextable<-ftable(data[,c("female","nas135")])
NaSextable

OddRatio<-(NaSextable[1,1]*NaSextable[2,2])/(NaSextable[1,2]*NaSextable[2,1])
ORVariance<-sum(1/c(NaSextable[1,],NaSextable[2,]))
```  
  
The Odds ratio is `r OddRatio` and the log of odd ratio is `r log(OddRatio)`  
The Variance is `r ORVariance` and the Std. Deviation is `r sqrt(ORVariance)`  
  
To obtain the 95% confidence interval for the log of Odd Ratio we apply $log(Odd\,Ratio)\pm1.96*Std.\,Deviation$
  
Thus, the upper bound is `r log(OddRatio)+(1.96*sqrt(ORVariance))` and the lower bound is `r log(OddRatio)-(1.96*sqrt(ORVariance))`  
  
Exponentiating we get the limits of the Odd Ratio. Upper bound is `r OddRatio*exp(1.96*sqrt(ORVariance))` and the lower bound is `r OddRatio/exp(1.96*sqrt(ORVariance))`  
 
The odds of a female experiencing hyponatremia is 3.4 times greater than that of a male. The 95% Confidence interval for the odds ratio is (1.97, 5.89). Upon repeated sampling, 95% of confidence intervals constructed this way would cover the true population odds ratio.  

  
b. Perform a logistic regression analysis with R using nas135 as dependent variable and female as the only independent variable. Use the Likelihood Ratio test to assess the significance of the model. Is the model with female a better model than the naïve model?  
  
```{r,echo=FALSE}
#We use glm function in R to get estimated parameters
lrmodel<-glm(data=data,nas135~female,family="binomial")
summary(lrmodel)
```  
  
  
The p-value of the likelihood test comparing the model with female and the one without is `r pchisq(lrmodel$null.deviance-lrmodel$deviance,df=1,lower.tail=FALSE)` The model with female is significantly better than the naïve model.  
  

c. What is the naïve model? What is the probability of hyponatremia that this model
predict?  
  
The naïve model (excluding any independent variable) can be seen in the table below  
  
```{r,echo=FALSE}
#We create the contingency table with nas135 in the columns and female in the rows
Natable<-ftable(data[,c("nas135")])
Natable
```  
  
The naïve model predicts a $\frac{62}{426+62}=$ `r (Natable[2]/sum(Natable))*100`% probability of hyponatremia for every subject in the study.  
  
  
d. Run a logistic regression analyses with no independent variables. Transform the
coefficient obtained from this model into a probability.  
  
```{r,echo=FALSE}
#We create the contingency table with nas135 in the columns and female in the rows
lrmodel0<-glm(data=data,nas135~1,family="binomial")
summary(lrmodel0)
```  
  
The odds is $Odds=\frac{Prob_{x=1}}{1-Prob_{x=1}}$. Manipulating that equation we can get that $Prob_{x=1}=\frac{Odds}{1+Odds}$  
  
The logit is the log(Odds), exponentiating we can get the probability from equation above $Prob_{x=1}=\frac{e^{logit}}{1+e^{logit}}$  
  
Using the value estimated in our model the probaility is `r (exp(lrmodel0$coeff))/(1+exp(lrmodel0$coeff))`  
  
  
e. Using the model with female as independent variable, compute the estimated
probability of hyponatremia per males and females. Write down the equation for the
logit.  
  
The model including female as independent variable is  
  
```{r,echo=FALSE}
#We create the contingency table with nas135 in the columns and female in the rows
lrmodel1<-glm(data=data,nas135~female,family="binomial")
summary(lrmodel1)
```  
  
The logit is $g(female)=\beta_0+\beta_1female$ 
  
Using the formula above, the probability of hyponatremia for a female is $Prob_{nas135=1}=\frac{e^{\beta_0+\beta_1}}{1+e^{\beta_0+\beta_1}}$= `r (exp(sum(lrmodel1$coeff)))/(1+exp(sum(lrmodel1$coeff)))`  
  
  
the probability of hyponatremia for a male is $Prob_{nas135=1}=\frac{e^{\beta_0}}{1+e^{\beta_0}}$= `r (exp(lrmodel1$coeff[1]))/(1+exp(lrmodel1$coeff[1]))` 
  
f. Use the Wald test to assess the significance of the coefficient for female.  
  
The Wald test is given in  the summary of the model above.  
  
  
g. Fit a model with runtime as the only independent variable. Assess the significance of the model.  
  
```{r,echo=FALSE}
#We create the contingency table with nas135 in the columns and female in the rows
lrmodel2<-glm(data=data,nas135~runtime,family="binomial")
summary(lrmodel2)
```    
  
The p-value of the likelihood test comparing the model with runtime and the one without it is `r pchisq(lrmodel2$null.deviance-lrmodel2$deviance,df=1,lower.tail=FALSE)` The model with runtime is significantly better than the naïve model.  
  
h. Calculate the probability of hyponatremia of a runner who takes 4 hours (240 minutes) to complete the marathon.  
  
The value of the logit for a runner taking 4 hours is `r sum(lrmodel2$coeff*c(1,240))` and the probablity following the equation described above is `r exp(sum(lrmodel2$coeff*c(1,240)))/(1+exp(sum(lrmodel2$coeff*c(1,240))))`  
  
  
i. Fit a model with female and runtime as independent variables. Assess the
significance of the model. Which null hypothesis is tested?
	
```{r,echo=FALSE}
#We create the contingency table with nas135 in the columns and female in the rows
lrmodel3<-glm(data=data,nas135~runtime,family="binomial")
summary(lrmodel3)
```  
  

The p-value of the likelihood test comparing the model with female and runtime and the one without them is `r pchisq(lrmodel3$null.deviance-lrmodel3$deviance,df=2,lower.tail=FALSE)` The model with female is significantly better than the naïve model.    